---
output: html_document
execute: 
  freeze: auto
editor_options: 
  chunk_output_type: console
---
# PRSice-2 {.unnumbered}

[PRSice-2](https://choishingwan.github.io/PRSice/) implements a Pruning and thresholding model to construct polygenic risk scores (PRS). Pruning involves removing closely linked genetic variants to reduce redundancy and potential bias due to linkage disequilibrium, ensuring that the most independent and informative markers are used. Thresholding, on the other hand, selects variants based on their p-values from GWAS, allowing only those variants that meet a specified significance threshold to contribute to the PRS. 

Choi SW, and O’Reilly PF. "PRSice-2: Polygenic Risk Score Software for Biobank-Scale Data." [GigaScience 8, no. 7 (July 1, 2019)](https://doi.org/10.1093/gigascience/giz082).

## Estimate AD-PRS
Here we will estimate three Alzheimer's disease PRS using different base GWAS. 

- Kunkle et al 2019. Clinicaly diagnosed AD 
- Bellenguez et al 2022. Alzheimer's disease and related dementias
- Laket et al 2023. Multi-ancestry meta analysis (MAMA) of Alzheimer's disease and related dementias

We are constructing the PRS across nine different P-value thresholds (5e-8, 1e-7, 1e-6, 1e-5, 1e-4, 0.001, 0.01, 0.1, 0.5, 1) and apply a clumping process with a window size of 250kb and an r2 > 0.1.  

::: {.callout-warning}
It is crucial to ensure that both the base and target datasets use the same human genome build. To achieve this, we have used MungeSummstats to liftover all summary statistics to build 38.
:::


### AD-PRS (Kunkle 2019)

```{bash}
#| label: ad_prs
#| code-fold: true
#| code-summary: AD-PRS PRSice
#| warning: false
#| error: false
#| message: false
#| eval: false

Rscript bin/PRSice.R --dir . \
    --prsice bin/PRSice \
    --base work/summary_statistics/Kunkle2019load_b38.tsv.gz \
    --snp SNP \
    --chr CHR \
    --bp BP \
    --A1 A2 \
    --A2 A1 \
    --stat BETA \
    --pvalue P \
    --target work/habshd_gwas \
    --pheno work/habshd_gwas.pheno \
    --cov work/habshd_gwas.covar \
    --thread 1 \
    --clump-kb 250kb \
    --clump-p 1.000000 \
    --clump-r2 0.100000 \
    --bar-levels 5e-8,1e-7,1e-6,1e-5,1e-4,0.001,0.01,0.1,0.5,1 \
    --fastscore \
    --no-default \
    --binary-target F \
    --all-score \
    --out work/Kunkle

```

### ADRD-PRS (Bellenguez 2022)
```{bash}
#| label: adrd_prs
#| code-fold: true
#| code-summary: ADRD-PRS PRSice
#| warning: false
#| error: false
#| message: false
#| eval: false

Rscript bin/PRSice.R --dir . \
    --prsice bin/PRSice \
    --base work/summary_statistics/Bellenguez20202adrd_b38.tsv.gz \
    --snp SNP \
    --chr CHR \
    --bp BP \
    --A1 A2 \
    --A2 A1 \
    --stat BETA \
    --pvalue P \
    --target work/habshd_gwas \
    --pheno work/habshd_gwas.pheno \
    --cov work/habshd_gwas.covar \
    --thread 1 \
    --clump-kb 250kb \
    --clump-p 1.000000 \
    --clump-r2 0.100000 \
    --bar-levels 5e-8,1e-7,1e-6,1e-5,1e-4,0.001,0.01,0.1,0.5,1 \
    --fastscore \
    --no-default \
    --binary-target F \
    --all-score \
    --out work/Bellenguez

```

### MAMA-PRS (Lake 2023)
```{bash}
#| label: mama_prs
#| code-fold: true
#| code-summary: MAMA-PRS PRSice
#| warning: false
#| error: false
#| message: false
#| eval: false

Rscript bin/PRSice.R --dir . \
    --prsice bin/PRSice \
    --base work/summary_statistics/Lake2023adrd_b38.tsv.gz \
    --snp SNP \
    --chr CHR \
    --bp BP \
    --A1 A2 \
    --A2 A1 \
    --stat BETA \
    --pvalue P \
    --target work/habshd_gwas \
    --pheno work/habshd_gwas.pheno \
    --cov work/habshd_gwas.covar \
    --thread 1 \
    --clump-kb 250kb \
    --clump-p 1.000000 \
    --clump-r2 0.100000 \
    --bar-levels 5e-8,1e-7,1e-6,1e-5,1e-4,0.001,0.01,0.1,0.5,1 \
    --fastscore \
    --no-default \
    --binary-target F \
    --all-score \
    --out work/Lake

```

## HABS-HD 

We will now evaluate the association each AD-PRS with the Clinical Dementia Rating Scale and cognitive impairment.

```{r}
#| label: packages
#| code-fold: true
#| code-summary: Import Packages
#| warning: false
#| error: false
#| message: false
#| eval: false

library(tidyverse)
library(pROC)
# library(janitor)
# library(broom)
# library(performance)

```


```{r}
#| label: data
#| code-fold: true
#| code-summary: Import and munge data
#| warning: false
#| error: false
#| message: false
#| eval: false

# File paths
pheno_path = "work/habshd_pheno.csv"
pcs_path = 'work/habs_pca.tsv'
ad_prs_path = 'work/prsice/Kunkle.all_score'
# adrd_prs_path = 'work/prsice/Bellenguez.all_score'
# mama_prs_path = 'work/prsice/Lake.all_score'

## HABS-HD Phenotypes
habshd <- read_csv(pheno_path) %>% distinct(med_id, .keep_all = T)

## Genetic Ancestry & PCs
pcs <- read_tsv(pcs_path) %>%
  filter(superpop == "HABS-HD") %>%
  select(-Population, -cohort, -superpop, -FID) %>%
  mutate(IID = as.numeric(IID)) %>%
  rename(superpop = superpop_infered) %>%
  filter(superpop != 'EAS')

## PRS  
prs <- read_table(ad_prs_path) %>%
  janitor::clean_names() %>%
  select(-fid)

## Merge datasets
dat <- habshd %>%
  left_join(pcs, by = c('med_id' = 'IID')) %>%
  left_join(prs, by = c('med_id' = 'iid')) %>%
  filter(!is.na(pt_5e_08) & !is.na(superpop)) %>%
  mutate(
    z_prs = scale(pt_5e_08)[,1],
    # z_prs = scale(pt_0_1)[,1],
    race = fct_relevel(race, "NHW"), 
    superpop = fct_relevel(superpop, "EUR"), 
    dx = fct_recode(as.factor(cdx_cog), 'ctrl' = '0', 'case' = '1', 'case' = '2'), 
    dx = fct_relevel(dx, 'ctrl'), 
    apoe = fct_recode(apoe4_snp, 
                      'e2+' = 'E2E2', 'e2+' = 'E2E3',  
                      'e4+' = 'E3E4', 'e4+' = 'E2E4', 'e4+' = 'E4E4', 
                      'e3/e3' = 'E3E3'
                      ), 
    apoe = fct_relevel(apoe, 'e3/e3')
  ) 

```

### PRS Distribution 

- What is the distribution of the AD-PRS across cases and controls? 
- What is the relationship between AD-PRS and CDR

```{r}
#| label: PRS
#| code-fold: true
#| code-summary: PRS
#| warning: false
#| error: false
#| message: false
#| eval: false
#| echo: true

## Violin Plots - DX
ggplot(dat, aes(x = dx, y = z_prs, fill = dx)) + 
  geom_violin() + 
  geom_boxplot(width = 0.2, outliers = FALSE, fill = 'white') +
  theme_bw()

## Density Plots - DX
ggplot(dat, aes(x = z_prs, fill = dx)) + 
  geom_density(alpha = 0.5) + 
  theme_bw()

## Scatter plot - CDR
ggplot(dat, aes(x = cdr_sum, y = z_prs)) + 
  geom_point() + 
  geom_smooth(method = 'lm') + 
  theme_bw()

## Violin Plots - super_pop
ggplot(dat, aes(x = superpop, y = z_prs, fill = superpop)) + 
  geom_violin() + 
  geom_boxplot(width = 0.2, outliers = FALSE, fill = 'white') +
  theme_bw()

```

### Predictive ability 
- What is the association of the AD-PRS with cognitive impairment


```{r}
#| label: OR
#| code-fold: true
#| code-summary: OR
#| warning: false
#| error: false
#| message: false
#| eval: false
#| echo: true

reduced_mod <- glm(dx ~ age + id_gender + PC1 + PC2 + PC3 + PC4, 
    data = dat, family = 'binomial')

full_mod <- glm(dx ~ z_prs + age + id_gender + PC1 + PC2 + PC3 + PC4, 
    data = dat, family = 'binomial')

broom::tidy(full_mod, exponentiate = T, conf.int = T)
```

### Predictive accuracy 
- What is the R2 of the reduced model containing only covariates (age, sex and PC1-4)?
- What is the R2 of the full model including the AD-PRS (age, sex and PC1-4)?

```{r}
#| label: R2
#| code-fold: true
#| code-summary: R2
#| warning: false
#| error: false
#| message: false
#| eval: false
#| echo: true

reduced_r2 <- performance::r2_nagelkerke(reduced_mod)
full_r2 <- performance::r2_nagelkerke(full_mod)

tribble(
  ~reduced, ~full, ~diff,
  reduced_r2, full_r2, full_r2 - reduced_r2
) 
```

### Discrimination 
- What is the AUC of the reduced model containing only covariates (age, sex and PC1-4)?
- What is the AUC of the full model including the AD-PRS (age, sex and PC1-4)?

```{r}
#| label: AUC
#| code-fold: true
#| code-summary: AUC
#| warning: false
#| error: false
#| message: false
#| eval: false
#| echo: true

# Predict probabilities
probabilities_full <- predict(full_mod, type = "response")
probabilities_reduced <- predict(full_mod, type = "response")

# Calculate AUC
roc_curve_full <- roc(response = dat$dx, predictor = probabilities_full)
roc_curve_reduced <- roc(response = dat$dx, predictor = probabilities_reduced)

auc(roc_curve_full)
auc(roc_curve_reduced)

ggroc(roc_curve_full) + 
  geom_abline(slope = 1, intercept = 1, linetype = 2) + 
  theme_bw()
```

### Calibration 
```{r}
#| label: calibration
#| code-fold: true
#| code-summary: Calibration
#| warning: false
#| error: false
#| message: false
#| eval: false
#| echo: true

probs <- dat %>% 
  select(med_id, dx) %>%
  mutate(
    predicted = probabilities_full
  ) %>%
  mutate(prob_bin = cut(predicted, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE))

cal_plot_breaks(probs, dx, predicted)

dat2$predicted_probs <- predict(full_mod, type = "response")

dat2 <- probs  %>%
  group_by(prob_bin) %>%
  summarise(observed_mean = mean(dx),
            # predicted_mean = mean(predicted),
            .groups = 'drop')

## 

glm(dx ~ z_prs + age + id_gender + id_education + PC1 + PC2 + PC3 + PC4, 
                data = dat %>% filter(race == 'Hispanic'), family = 'binomial') %>%
  broom::tidy()

lm(cdr_sum ~ z_prs + age + id_gender + PC1 + PC2 + PC3 + PC4, 
   data = dat %>% filter(superpop == 'AFR')) %>%
  broom::tidy()


```


```{r}
#| label: test1
#| code-fold: true
#| code-summary: Import Packages
#| warning: false
#| error: false
#| message: false
#| eval: false
#| echo: false

lm(cdr_sum ~ z_prs + age + id_gender + race + PC1 + PC2 + PC3 + PC4, 
   data = dat) %>%
  broom::tidy()
```


```{r}
#| label: test2
#| code-fold: true
#| code-summary: Import Packages
#| warning: false
#| error: false
#| message: false
#| eval: false
#| echo: false

full_mod <- glm(dx ~ z_prs + age + id_gender + id_education + race + PC1 + PC2 + PC3 + PC4, 
    data = dat, family = 'binomial')

broom::tidy(full_mod)

nagelkerke_r2 <- performance::r2_nagelkerke(full_mod)
print(nagelkerke_r2)

library(pROC)

# Predict probabilities
probabilities <- predict(full_mod, type = "response")

# Calculate AUC
roc_curve <- roc(response = dat$dx, predictor = probabilities)
auc_value <- auc(roc_curve)
print(auc_value)

## Calibration 
dat2 <- dat
dat2$predicted_probs <- predict(full_mod, type = "response")

dat2 <- dat2 %>%
  mutate(prob_bin = cut(predicted_probs, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)) %>%
  group_by(prob_bin) %>%
  summarise(observed_mean = mean(dx),
            predicted_mean = mean(predicted_probs),
            .groups = 'drop')

## 

glm(dx ~ z_prs + age + id_gender + id_education + PC1 + PC2 + PC3 + PC4, 
                data = dat %>% filter(race == 'Hispanic'), family = 'binomial') %>%
  broom::tidy()

lm(cdr_sum ~ z_prs + age + id_gender + PC1 + PC2 + PC3 + PC4, 
   data = dat %>% filter(superpop == 'AFR')) %>%
  broom::tidy()


```



























